Verification of the CVM algorithm with a Functional Probabilistic Invariant

### 2025-03-04 paper/main.tex
Paper: Minor edits.   * Use central place for LOCs.
minor edits
fix bib except math
add more line count to show benefit of invariant
Edit intro slightly.   - Move line count for neg assoc library to section 6.
assorted minor fixes
main - latest line count, bib url
intro - minor
sec 2 - fix probabilities
sec 3 - minor on role of q
sec 4 - minor edits
sec 5 - minor edits
sec 6 - minor edits
sec 7 - mention relational reasoning, fix small bug

\documentclass[a4paper,UKenglish,cleveref, autoref, thm-restate]{lipics-v2021}
%This is a template for producing LIPIcs articles.
%See lipics-v2021-authors-guidelines.pdf for further information.
%for A4 paper format use option "a4paper", for US-letter use option "letterpaper"
%for british hyphenation rules use option "UKenglish", for american hyphenation rules use option "USenglish"
%for section-numbered lemmas etc., use "numberwithinsect"
%for enabling cleveref support, use "cleveref"
%for enabling autoref support, use "autoref"
%for anonymousing the authors (e.g. for double-blind review), add "anonymous"
%for enabling thm-restate support, use "thm-restate"
%for enabling a two-column layout for the author/affilation part (only applicable for > 6 authors), use "authorcolumns"
%for producing a PDF according the PDF/A standard, add "pdfa"

@@

\usepackage{tikz}
\usetikzlibrary{calc}
\usetikzlibrary{shapes.geometric}
\usetikzlibrary{decorations.pathreplacing,calligraphy}
\newcommand*\circled[1]{\tikz[baseline=(char.base)]{
            \node[shape=circle,draw, minimum size=3.5mm,inner sep=0.5pt] (char) {#1};}}

@@

@@

\author{\textcolor{red}{Anounymous authors}}{\textcolor{red}{Anonymous affiliations}}{}{}{}

\authorrunning{\textcolor{red}{Anonymous}} %TODO mandatory. First: Use abbreviated first/middle names. Second (only in severe cases): Use first author plus 'et al.'

\Copyright{\textcolor{red}{Anonymous}} %TODO mandatory, please use full first names. LIPIcs license is "CC-BY";  http://creativecommons.org/licenses/by/3.0/

>>>>

\RequirePackage[hyphens]{url}
\documentclass[a4paper,UKenglish,cleveref, autoref, thm-restate]{lipics-v2021}
%This is a template for producing LIPIcs articles.
%See lipics-v2021-authors-guidelines.pdf for further information.
%for A4 paper format use option "a4paper", for US-letter use option "letterpaper"
%for british hyphenation rules use option "UKenglish", for american hyphenation rules use option "USenglish"
%for section-numbered lemmas etc., use "numberwithinsect"
%for enabling cleveref support, use "cleveref"
%for enabling autoref support, use "autoref"
%for anonymousing the authors (e.g. for double-blind review), add "anonymous"
%for enabling thm-restate support, use "thm-restate"
%for enabling a two-column layout for the author/affilation part (only applicable for > 6 authors), use "authorcolumns"
%for producing a PDF according the PDF/A standard, add "pdfa"

@@

\usepackage{tikz}
\usetikzlibrary{arrows.meta}
\usetikzlibrary{calc}
\usetikzlibrary{shapes.geometric}
\usetikzlibrary{decorations.pathreplacing,calligraphy}
\newcommand*\circled[1]{\tikz[baseline=(char.base)]{
            \node[shape=circle,draw, minimum size=3.5mm,inner sep=0.5pt] (char) {#1};}}

@@

\newcommand\locnew{1003}
\newcommand\locold{2630}

@@

\author{\textcolor{red}{Anonymous author(s)}}{\textcolor{red}{Anonymous affiliation(s)}}{}{}{}

\authorrunning{\textcolor{red}{Anonymous author(s)}} %TODO mandatory. First: Use abbreviated first/middle names. Second (only in severe cases): Use first author plus 'et al.'

\Copyright{\textcolor{red}{Anonymous author(s)}} %TODO mandatory, please use full first names. LIPIcs license is "CC-BY";  http://creativecommons.org/licenses/by/3.0/

### 2025-03-04 paper/sec/formalization_neg_dep.tex

Another key result (shown below) used for the verification of our CVM variant is the negative-associativity of the indicator functions of random $k$-subsets of a finite set $S$ (with cardinality greater than $k$).

>>>>

Another key result (shown below) used for the verification of our CVM variant is the negative-associativity of the indicator functions of random $k$-subsets of a finite set $S$ (with cardinality greater than or equal to $k$).

### 2025-03-04 paper/sec/negdep.tex

An interesting consequence of our invariant-based approach, is that it allowed us to devise and verify a refined version of the CVM algorithm, that is both total and unbiased.

@@

\begin{algorithm}[t!]
	\caption{New unbiased and total CVM algorithm variant.}\label{alg:cvm_new}
	\begin{algorithmic}[1]
  \Require Stream elements $a_1,\dots,a_l$, $0 < \varepsilon$, $0 < \delta < 1$.
  \Ensure A cardinality estimate $R$ for set $A = \{ a_1,\dots,a_l \}$ s.t. $\prob \left( |R - |A| | > \varepsilon |A| \right) \leq \delta$
  \State $\chi \gets \{\}, p \gets 1, n = \ceil*{\frac{12}{\varepsilon^2} \ln{(\frac{3l}{\delta})} }$, $\frac{1}{2} \leq f < 1$, s.t., $nf$ integer
  \For{$i \gets 1$ to $l$}
    \State $b \getsr \Ber(p)$ \Comment insert $a_i$ with probability $p$ (and remove it otherwise)
    \If{$b$}
      \State $\chi \gets \chi \cup \{a_i\}$
    \Else
      \State $\chi \gets \chi - \{a_i\}$
    \EndIf
    \If{$|\chi| = n$} \Comment if buffer $\chi$ is full
      \State $\chi \getsr \mathrm{subsample}(\chi)$ \Comment select a random $nf$-subset of $\chi$
      \State $p \gets pf$
    \EndIf
  \EndFor
  \State \Return $\frac{|\chi|}{p}$ \Comment estimate cardinality of $A$
  \end{algorithmic}
\end{algorithm}%

@@

Motivated by this, our new variant is shown in \cref{alg:cvm_new}.
For the subsampling step, instead of keeping each element of $\chi$ with probability $\frac{1}{2}$, we instead pick a uniform random $nf$-subset, where $\frac{1}{2} \leq f < 1$ and such that $nf$ is an integer.\footnote{It is possible to choose $f = \frac{n-1}{n}$, i.e., discarding a random element from $\chi$ in the subsampling step.}
Since this new subsampling step always reduces the size of $\chi$, the variant never returns $\bot$, i.e., it is \emph{total}.
The invariant-based approach allows us to show that the algorithm is probably-approximately correct and also \emph{unbiased}, i.e., the expectation of the result is exactly $|A|$.
These depend crucially on establishing~\cref{i:subsample_condition} for the new subsampler, for which we need a new concept.

@@

\begin{definition}[Negative Association]\label{def:neg_assoc}
A set of random variables $X_1,\dots,X_n: \Omega \rightarrow \mathbb R$ is negatively associated if, for all non-decreasing functions $f,g: \mathbb R^n \rightarrow \mathbb R$, which depend on disjoint sets of the variables, i.e., $\mathrm{dep}(f) \cap \mathrm{dep}(g) = \emptyset$, the following inequality holds.
\[
\expect [f(X_1,\ldots,X_n) g(X_1,\ldots,X_n)] \leq \expect [f(X_1,\ldots,X_n)] \expect [g(X_1,\ldots,X_n)] \textrm{.}
\]
\end{definition}

@@

The case for non-decreasing functions of the above lemma is pointed out by Joag-Dev and Proschan~\cite[P.2]{joagdev1983}.
The reason for our interest in this lemma stems from the fact that indicator variables of random $m$-subsets are negatively associated.
This is a consequence of the fact that permutation distributions are negatively-associated~\cite[Th. 2.11]{joagdev1983}.
Thus, for the new subsampling step in Line 9 of~\cref{alg:cvm_new}, we can derive using \cref{le:neg_assoc_prod}:
\begin{equation}\label{eq:subsample_with_n_subsets}
  \int_{\mathrm{subsample}(\chi)} \prod_{s \in S} g(\indicat(s \in \tau))\,d\tau \leq
  \prod_{s \in S} \int_{\mathrm{subsample}(\chi)} g(\indicat(s \in \tau))\,d\tau = \prod_{s \in S}  \expect_{\Ber(f)}[g] \textrm{.}
\end{equation}
for any non-negative $g$ and $S \subseteq \chi$.
Note that the domain of $g$ has two values, so it is either non-increasing or non-decreasing.
Also, if $S$ is a singleton, the inequality becomes an equality.
With this ingredient, we can conclude that our results about the original algorithm we derived in the previous section also hold for our new alternative algorithm (\cref{alg:cvm_new}).
% Note: After reading this section. It stops too abrubtly at the previous sentence, without establishing what we have actually shown.

>>>>

An interesting consequence of our invariant-based approach is that it allowed us to devise and verify a refined version of the CVM algorithm that is both total and unbiased.

@@

\begin{algorithm}[t!]
	\caption{New total and unbiased CVM algorithm variant.}\label{alg:cvm_new}
	\begin{algorithmic}[1]
  \Require Stream elements $a_1,\dots,a_l$, $0 < \varepsilon$, $0 < \delta < 1$.
  \Ensure A cardinality estimate $R$ for set $A = \{ a_1,\dots,a_l \}$ s.t. $\prob \left( |R - |A| | > \varepsilon |A| \right) \leq \delta$
  \State $\chi \gets \{\}, p \gets 1, n = \ceil*{\frac{12}{\varepsilon^2} \ln{(\frac{3l}{\delta})} }$, $\frac{1}{2} \leq f < 1$, s.t., $nf$ integer
  \For{$i \gets 1$ to $l$}
    \State $b \getsr \Ber(p)$ \Comment insert $a_i$ with probability $p$ (and remove it otherwise)
    \If{$b$}
      \State $\chi \gets \chi \cup \{a_i\}$
    \Else
      \State $\chi \gets \chi - \{a_i\}$
    \EndIf
    \If{$|\chi| = n$} \Comment if buffer $\chi$ is full
      \State $\chi \getsr \mathrm{subsample}(\chi)$ \Comment select a random $nf$-subset of $\chi$
      \State $p \gets pf$
    \EndIf
  \EndFor
  \State \Return $\frac{|\chi|}{p}$ \Comment estimate cardinality of $A$
  \end{algorithmic}
\end{algorithm}%

@@

Motivated by this observation, our new variant is shown in \cref{alg:cvm_new}.
For the subsampling step, instead of keeping each element of $\chi$ with probability $\frac{1}{2}$, we instead pick a uniform random $nf$-subset of $\chi$, where $\frac{1}{2} \leq f < 1$ and such that $nf$ is an integer.
For example, it is possible to choose $f = \frac{n-1}{n}$, i.e., discarding a random element from $\chi$ in the subsampling step.
Since this new subsampling step always reduces the size of $\chi$, the variant is \emph{total} (never returns $\bot$).
The invariant-based approach allows us to show that the algorithm is probably-approximately correct and also \emph{unbiased}, i.e., the expectation of the result is exactly $|A|$.
These depend crucially on establishing~\cref{i:subsample_condition} for the new subsampler, for which we need a new concept.

@@

\begin{definition}[Negative Association]\label{def:neg_assoc}
A set of random variables $X_1,\dots,X_n: \Omega \rightarrow \mathbb R$ is negatively associated if, for all non-decreasing functions $f,g: \mathbb R^n \rightarrow \mathbb R$, which depend on disjoint sets of the variables, i.e., $\mathrm{dep}(f) \cap \mathrm{dep}(g) = \emptyset$, the following inequality holds:
\[
\expect [f(X_1,\ldots,X_n) g(X_1,\ldots,X_n)] \leq \expect [f(X_1,\ldots,X_n)] \expect [g(X_1,\ldots,X_n)] \textrm{.}
\]
\end{definition}

@@

The case for non-decreasing functions of the above lemma is pointed out by Joag-Dev and Proschan~\cite[P.2]{joagdev1983}.
The reason for our interest in this lemma stems from the fact that indicator variables of random $m$-subsets are negatively associated.
This is a consequence of the fact that permutation distributions are negatively-associated~\cite[Th. 2.11]{joagdev1983}.
Thus, for the new subsampling step in Line 9 of~\cref{alg:cvm_new}, we can derive using \cref{le:neg_assoc_prod}:
\begin{equation}\label{eq:subsample_with_n_subsets}
  \int_{\mathrm{subsample}(\chi)} \prod_{s \in S} g(\indicat(s \in \tau))\,d\tau \leq
  \prod_{s \in S} \int_{\mathrm{subsample}(\chi)} g(\indicat(s \in \tau))\,d\tau = \prod_{s \in S}  \expect_{\Ber(f)}[g] \textrm{.}
\end{equation}
for any non-negative $g$ and $S \subseteq \chi$.
Note that the domain of $g$ has two values, so it is either non-increasing or non-decreasing.
Also, if $S$ is a singleton, the inequality becomes an equality.
With this ingredient, we can conclude that our results about the original algorithm derived in the previous section also hold for our new variant (\cref{alg:cvm_new}).
% Note: After reading this section. It stops too abrubtly at the previous sentence, without establishing what we have actually shown.

### 2025-03-04 paper/sec/original_proof.tex

It is easy to show for \circled{C} that the coin flips are independent, and that the set of elements in $\chi$ in any state are exactly those stream elements for which the first $k$ entries of their associated coin flips are $1$.
The final random process $\circled{D}$ directly computes the final set of elements in $\chi$ after the stream, taking $K$ as a fixed parameter; one relates \circled{C} to \circled{D} by:
\[ \prob_{\circled{C}}(k = K \land \chi = X)  \leq \prob_{\circled{D}_K}(\chi = X)\]
for fixed values of $K$ and $X$.
To see how tail bounds can be derived from this inequality, note that it is possible to union bound the probability that the estimate exceeds the desired interval, with the probability of the event in conjunction with specific values $k = K$.
The latter can be bounded using a Chernoff bound for the probability that the number of stream elements whose associated coin flips start with $K$ $1$s is outside $2^{-K} |A| (1 \pm \varepsilon)$.
This is explained in more detail by Chakraborty et al.~\cite{chakraborty2023}.

>>>>

It is easy to show for \circled{C} that the coin flips are independent, and that the set of elements in $\chi$ in any state are exactly those stream elements for which the first $k$ entries of their associated coin flips are $1$.
The final random process $\circled{D}$ directly computes the final set of elements in $\chi$ after the stream, taking $K$ as a fixed parameter; one relates \circled{C} to \circled{D} by:
\[ \prob_{\circled{C}}(k = K \land \chi = X)  \leq \prob_{\circled{D}_K}(\chi = X)\]
for fixed values of $K$ and $X$.
To see how tail bounds can be derived from this inequality, let us first consider the failure event where the algorithm \circled{C}'s estimate exceeds the desired estimation interval and it ends with some fixed value $k = K$.
Using \circled{D}, this can be bounded using a Chernoff bound for the probability that the number of stream elements whose associated coin flips start with $K$ $1$s is outside $2^{-K} |A| (1 \pm \varepsilon)$.
Now, we can take a union bound over all the possible values $K$ to establish a global bound for the failure event in \circled{C}.
This is explained in more detail by Chakraborty et al.~\cite{chakraborty2023}.
